A Constraint Solving Approach to Parikh Images of Regular Languages

### 2024-03-07 introduction.tex
more

The problem can be solved using the classical construction of Parikh
images~\cite{muscholl-linear,generate-parikh-image} by first computing
the product of finite-state automata accepting the
languages~$\Language_1, \ldots, \Language_n$, then extracting a
Presburger arithmetic formula~$\psi$ describing the Parikh image of the product
automaton, and finally checking the satisfiability of $\phi \wedge \psi$.

@@

Secondly, the constructed Presburger arithmetic formula~$\psi$ contains a linear
number of existential quantifiers in the size of the product
automaton, as well as complex Boolean structure, which is needed to
express the connectedness of paths considered in the construction.
Solving formulas of this kind generally tends to be taxing for
solvers~\cite{ostrich-plus}; in the present case, since the product
automaton itself is exponentially big, also formula~$\psi$ has
exponential size in the number of considered regular languages, and
can easily become too complex for today's Presburger arithmetic
solvers to handle.

@@

Addressing these concerns, we have developed a calculus for Parikh images of
intersections of regular languages that we call \Calculus{}. It allows us to
interleave the computation of arbitrarily deep products of automata with the
product's Parikh image, and is generalised to an arbitrary homomorphism over
automata labels, including string length. This enables us to let both
calculations inform each other, eliminating unnecessary work, and pruning the
size of the partial products considered in the computation for a smaller memory
footprint. Moreover, our approach allows us to do this precisely, without the expensive
fallbacks when approximations fail as in the case of~\cite{approximate-parikh}.

>>>>

The problem can be solved using the classical construction of Parikh
images~\cite{muscholl-linear,generate-parikh-image} by first computing
the product of finite-state automata accepting the
languages~$\Language_1, \ldots, \Language_n$, then extracting a
Presburger arithmetic formula~$\rho$ describing the Parikh image of the product
automaton, and finally checking the satisfiability of $\phi \wedge \rho$.

@@

Secondly, the constructed Presburger arithmetic formula~$\rho$ contains a linear
number of existential quantifiers in the size of the product
automaton, as well as complex Boolean structure, which is needed to
express the connectedness of paths considered in the construction.
Solving formulas of this kind generally tends to be taxing for
solvers~\cite{ostrich-plus}; in the present case, since the product
automaton itself is exponentially big, also formula~$\rho$ has
exponential size in the number of considered regular languages, and
can easily become too complex for today's Presburger arithmetic
solvers to handle.

@@

Addressing these concerns, we have developed a calculus for Parikh images of
intersections of regular languages that we call \Calculus{}. It allows us to
interleave the computation of product automata with the derivation of the
Parikh image. This enables the two
calculations to inform each other, eliminating unnecessary work, and pruning the
size of the partial products considered in the computation for a smaller memory
footprint.
Moreover, our approach allows us to do this precisely, without the expensive
fallbacks when approximations fail as in the case of~\cite{approximate-parikh}.
%
Our calculus is applicable not only to Parikh images,
but can be used for reasoning about the image of regular languages under
arbitrary homomorphisms into commutative monoids. Besides Parikh images,
this includes the length abstraction of regular languages commonly used in
string solvers.
